---
title: "Run UTPC on the big datasets"
output: html_notebook
---


```{r setup}
library(tidyverse)
library(magrittr)
library(readxl)
library(patchwork)
library(plotly)
library(ggforce)
```



## Import the data


```{r}

big_dd <- read.csv("data/thermal_performance_datasets.csv") %>%
  rename(type = `id`) %>%
  select(-citation)

```





### Fit our own parameters using UTPC method

We optimise on the least squares of the rescaled data to the universal thermal performance curve using the L-BFSG-U approach to bound the parameters. 

```{r}

# define our fitness function which is the sum of squares of 
# the predicted - observed
tpc_fit <- function(x, pf, ta){
  
  # extract the parameters
  yopt  <- x[1] # the value of y at the optimum
  Topt  <- x[2] # the optimal temperature
  Delta <- x[3] # the difference between Tc-Topt
  Tc    <- Topt + Delta # calculate Tc
  
  # sum of squares
  sum(((yopt * exp( (ta - Topt) / Delta) * (1 - (ta - Topt)/Delta)) - pf) ^ 2)
  
  
}

# a function to find sensible initial values for the optimiser
get_inits <- function(x){
  
  # initialise the initial parameter value vector
  inits <- c(NA, NA, NA)
  
  # find the largest performance
  idx_max_y <- which.max(x$pf)
  
  # the largest y value is a good guess for yopt
  inits[1] <- x$pf[idx_max_y]
  
  # the corresponding temperature is a good guess for Topt
  inits[2] <- x$ta[idx_max_y]
  
  # use the largest T value as a simple guess for Delta
  inits[3] <-tail(x$ta[is.finite(x$ta)],1) - inits[2]
  
  # return the initial values
  return(inits)
}


# a wrapper function to do the fit and return a data.frame
# The object x coming in is a data.frame of the same structure
# as the full dataset. The output is a tibble of the same structure
# with columns appended for estimates of the universal TPC with 
# prefix "u_"
do_tpc_fit <- function(x){
  
  # call the optimiser
  fit <- optim(get_inits(x), 
               tpc_fit,
               pf = x$pf, 
               ta = x$ta,
               method = "L-BFGS-B",
               lower = c(0.2, 10^-3, 1), 
               upper = c(10, 100, Inf))
  
  # return a tibble (data.frame)
  return(x %<>% mutate(u_yopt  = fit$par[1],
                       u_Topt  = fit$par[2],
                       u_Delta = fit$par[3],
                       u_Tc    = fit$par[2] + fit$par[3], 
                       SS = fit$value)
         )

  
}


# this function takes a fitted UTPC model and scales the data to the 
# UTPC. Returns a data.frame the same as the original data.frame
# with new columns for 
# scaled performance pf_utpc 
# and scaled temperature ta_utpc
data2utpc <- function(dd){
  dd %<>% mutate(pf_utpc = pf / u_yopt,
                ta_utpc = (ta - u_Topt) / u_Delta
                )
  return(dd)
}

# This function scales the UTPC curve to the original data based on 
# the fitted UTPC model. Returns a new data.frame with n values of 
# the UTPC function evaluated and rescaled to the data provied in dd.
# Performance can be adjusted according by subtracting the value given in 
# p_adj, and would typically be used to allow for negative performance values
# in the data. 
utpc2data <- function(dd, y_adj = 0, n = 100){
  UTPC <- data.frame(x = seq(-4, 1, length.out = n)) %>%
  mutate(y = (1-x)*exp(x)) %>%
  mutate(ta = (x * dd$u_Delta[1]) + dd$u_Topt[1],
         pf = (y * dd$u_yopt[1] - y_adj),
         type = dd$type[1],
         pf_max = dd$pf_max[1]
)
  return(UTPC)
}
```

### Map over all the data

```{r}
# ph <- "Proteobacteria"

# big_dd <- big_dd %>%
#   filter(specific_trait_group == "population_growth_rate")

# rescale performance data to be numbers closer to 1
big_dd <- big_dd %>% group_by(type) %>% 
  mutate(ta = temperature,
         pf_max = max(trait_value),
         pf = trait_value / pf_max)


# split the data on "code" variable and map our optimiser over all
# the possibly() function acts as an error catcher, and then rebind it 
# into a single tibble again on completion.
big_dd %<>% 
  split(.$type) %>% 
  map(~possibly(do_tpc_fit, otherwise = 'error here')(.x)) %>%
  bind_rows()

## save this object to file
save(big_dd, 
     file = "data/combined_dat_params.rda", 
     compress = "xz")

## write to file
# write.csv(combined_dat_params, 
#           file = "data_for_JF.csv", 
#           row.names = FALSE)

```


## Visualise the UTPC on all data

```{r}

dd_utpc <- data2utpc(big_dd)

aj <- big_dd %>% split(.$type) %>% map(utpc2data) %>% bind_rows()

```

separate panels by phylum

```{r}

# dd_utpc %>% split(.$phylum) %>% map()

```


```{r}

gg1 <- ggplot(data = dd_utpc, mapping = aes(x = ta_utpc, 
                                       y = pf_utpc)) + 
  geom_point() + 
  facet_wrap(vars(phylum), 
             scales = "free_x") +
  # ylim(0,1) + 
  geom_function(fun = function(x) pmax((1-x) * exp(x), 0),
                col = "red")
  # ylim(0, 10)

print(gg1)
```


## Plot the UTPCs excluding data and models that do not fit well


UTPC transformed data. 

NB plot only curves which have 

```{r}
g_focus_utpc <- ggplot(data = dd_utpc %>% 
                         filter(u_yopt <= 2) %>%
                      mutate(type = factor(type)), 
                  mapping = aes(x = ta_utpc, y = pf_utpc)) + 
  geom_point(size = 1, alpha = 0.2) + 
  scale_color_viridis_d(guide="none") +
  # scale_color_viridis_d() + 
  geom_function(fun = function(x) pmax((1-x) * exp(x), 0),
                col = "white", linewidth = 1, xlim = c(-6, 1)) + 
  ylim(-0.5, 5) +
  xlim(-6, 1.1) + 
  theme_classic() + 
  xlab(expression(italic("(T"["trial"]*" - T"["opt"]*") / (T"["c"]*" - T"["opt"]*")"))) +
  ylab( expression("Pf / Pf"["max"]*"") )
  

print(g_focus_utpc)
```
## Find the problem fits

```{r}
dd_alg <- dd_utpc %>% filter(u_yopt > 2)

prbs <- unique(dd_alg$type)       

prbs <- sort(unique(prbs))

dd_prbs <- dd_utpc[dd_utpc$type %in% prbs,]


utpc_prbs <- dd_prbs %>% split(.$type) %>% map(utpc2data) %>% bind_rows()


```

plot them

```{r}

tmp <- data.frame(u_Delta = 7, u_Topt = 42, u_yopt = 0.00073, 
                  type = -999)

tmp_utpc = utpc2data(tmp)


g_raw_prbs <- ggplot(data = dd_prbs %>% 
                        mutate(type = factor(type)), 
                  mapping = aes(x = ta, y = pf / pf_max)) + 
  geom_point() + 
  geom_line() + 
  geom_line(data = utpc_prbs %>%
              mutate(type = factor(type)), linetype = 2) +
  facet_wrap_paginate(vars(type), scales = "free", ncol = 5, nrow = 4, 
                      page = 1)
  # scale_color_viridis_d(guide="none")
  # geom_line(data = tmp_utpc, color = "black")
  
  

print(g_raw_prbs)

```

```{r}

dd_prbs %>% group_by(type) %>%
  summarise(u_Topt = u_Topt[1], u_yopt = u_yopt[1], pf_max = pf_max[1])

```

### Filter for the zeroes

```{r}
zero_fun <- function(x){ifelse(any(x == 0), 0, 1)}

zz_utpc$zeros <- ifelse(dd_utpc$ta_utpc >= -2 & 
                              dd_utpc$ta <= 0 & 
                              dd_utpc$pf_utpc <= 0.01, 1, 0)

zz_utpc <- zz_utpc %>% 
  group_by(type) %>%
  mutate(zero_test = zero_fun(zeros)) %>% 
  ungroup()

g_zeros_prbs <- ggplot(data = zz_utpc %>% 
                         filter(zero_test == 0) %>% 
                        mutate(type = factor(type)), 
                  mapping = aes(x = ta_utpc, y = pf_utpc, color = type)) + 
  geom_point() + 
  # geom_line(data = utpc_prbs %>%
  #             mutate(type = factor(type)), linetype = 2) +
  # facet_wrap(vars(type), scales = "free") 
  scale_color_viridis_d(guide="none") + 
  # geom_line(data = tmp_utpc, color = "black") 
  xlim(-6, 1.1)
  

print(g_zeros_prbs)

```

