---
title: "Import data, generate parameter estimates and merge with Rezende data"
output: html_notebook
---


```{r setup}
library(tidyverse)
library(magrittr)
library(readxl)
library(patchwork)
```

This file imports the three datasets from Rezende & Bozinovic 2019. This file outputs a tibble data object, saved as an `*.rda` file comprising the original data alongside parameter estimates from Rezende & Bozinovic and parameter estimates following our UTPC method. The output file is in `output/combined_dat_params.rda`.


## Import the data

A function to extract the data from the original files. 

```{r}

rearrange_data <- function(x){
  
  
  # x <- "data/Fitness_raw_data.xlsx"
  
  # read in as tab separated
  dd <- read_excel(x, col_names = FALSE)
  
  # convert to matrix
  dd <- as.matrix(dd)
  
  # take off the head of the matrix and keep it to extract later
  head_dd <- dd[1:3,]
  
  # delete the head
  dd <- dd[-(1:3),]
  
  ## swap out the "NA" character for actual NA
  dd[dd == "NA"] <- NA
  
  # coerce to numeric which annoyingly throws warnings for coercing the NAs
  dd <- apply(dd, 2, as.numeric)
  
  # prep a list in which to store it so we can grow it on the fly (yuck)
  out <- list()
  
  
  for (i in 2:ncol(dd)) {
    
    which_real <- which(!is.na(dd[, i]))
    
    out[[i-1]] <- tibble( ta   = dd[which_real, 1],
                          pf   = dd[which_real, i],
                          name = head_dd[2, i],
                          code = head_dd[3, i],
                          ref  = head_dd[1, i]
    )
    
  }
  
  return(bind_rows(out))
  
}


```


Import the data, combine the three and separate the name column in to the meta data.

```{r}

# read each of the tree data files using our data extraction helper function
# rearrange_data()

# Read in the photosynthesis data
pho <- rearrange_data("data/Biochemistry_raw_data.xlsx")

# read in the population growth data
pop <- rearrange_data("data/Fitness_raw_data.xlsx")

# read in the lizard running data
run <- rearrange_data("data/Running_raw_data.xlsx")

# bind all three of them together into a single data.frame object
all_dat <- bind_rows(list(pho = pho, pop = pop, run = run), .id = "type")

# separate the name column which has variable number of "_" on which to split
# I recover genus and species and dump the rest into variables.
all_dat %<>% separate(name, 
                      into = c("genus", "species", "v3", "v4", "v5", 
                               "v7", "v8"), 
                      sep = "_", 
                      fill = "right")


# import the parameter estimates from the Rezende & Bozinovic paper 
# and drop the code2 duplicate column
# which i have checked is the same as code.
model_params <- read_csv("data/All_parameters_data.csv") %>% 
  select(-code2) %>% mutate(type2 = type) %>% select(-type)

# merge the two datasets by experimental code identifier
combined_dat_params <- full_join(all_dat, model_params, by = "code") 
```

There are `r length(unique(combined_dat_params$code))` experiments in this this dataset which can be used to fit TPCs. These are split over three broad classes describing biochemical rates of photosynthesis in plants (n = `r length(unique(pho$code))`), population growth in insects (n = `r length(unique(pop$code))`) and running speeds of lizards (n = `r length(unique(run$code))`).





### Fit our own parameters using UTPC method

We optimise on the least squares of the rescaled data to the universal thermal performance curve using the L-BFSG-U approach to bound the parameters. 

```{r}

# define our fitness function which is the sum of squares of 
# the predicted - observed
tpc_fit_scale_shape <- function(x, pf, ta){
  
  # rescale the temperature data
  xx <- (ta - x[1]) / x[2]
  
  # calculate the UTPC
  fx <- (1-xx)*exp(xx)
  
  # compare the performance data with the scaled UTPC
  sum( ( pf - x[4] - x[3] * fx ) ^ 2 )
  
  
}

# a function to find sensible initial values for the optimiser
get_inits <- function(x){
  
  # initialise the initial parameter value vector
  inits <- c(NA, NA, NA, NA)
  
  # find the largest performance
  idx_max_y <- which.max(x$pf)
  
  # the corresponding temerature is a good guess for Topt
  inits[1] <- x$ta[idx_max_y]
  
  # use the largest T value - Topt as a simple guess for Delta
  inits[2] <- max(c( max(x$ta) - inits[1], 5) )
  
  # the largest y value is a good guess for yopt
  inits[3] <- x$pf[idx_max_y]
  
  # offset for y
  inits[4] <- 0
  
  # return the initial values
  return(inits)
}

# get lower value bounds based on initial parameter values
get_lwr <- function(x){
  c(0.1*x[1]-5, 0.1*x[2], 0.1*x[3], -0.2*x[3])
}

# get upper value bounds based on initial parameter values
get_upp <- function(x){
  c(10*x[1]+5, 10*x[2], 10*x[3], 0.2*x[3])
}


# a wrapper function to do the fit and return a data.frame
# The object x coming in is a data.frame of the same structure
# as the full dataset. The output is a tibble of the same structure
# with columns appended for estimates of the universal TPC with 
# prefix "u_"
do_tpc_fit <- function(x){
  
  x0 <- get_inits(x)
  
  # call the optimiser
  fit <- optim(x0, 
               tpc_fit_scale_shape,
               pf = x$pf, 
               ta = x$ta,
               method = "L-BFGS-B",
               lower = get_lwr(x0), 
               upper = get_upp(x0))
  
  # return a tibble (data.frame)
  return(x %<>% mutate(u_yopt  = NA,
                       u_Topt  = NA,
                       u_Delta = NA,
                       u_Tc    = NA,
                       x1 = fit$par[1], 
                       x2 = fit$par[2], 
                       x3 = fit$par[3], 
                       x4 = fit$par[4])
         )

  
}

```

### Map over all the data

```{r}
# split the data on "code" variable and map our optimiser over all
# the possibly() function acts as an error catcher, and then rebind it 
# into a single tibble again on completion.
combined_dat_params_2 <- combined_dat_params %>% 
  split(.$code) %>% 
  map(~possibly(do_tpc_fit, otherwise = 'error here')(.x)) %>%
  bind_rows()

## save this object to file
save(combined_dat_params_2, 
     file = "output/combined_dat_params_shape_and_scale.rda", 
     compress = "xz")


```


### References

Rezende, E. L. & Bozinovic, F. 2019. Thermal performance across levels of biological organization. Philosophical Transactions of the Royal Society B 374, 20180549.

## Visualise model Fit

```{r}

# Resecale the temperature and performance data according to the model fit
combined_dat_params_2 %<>% mutate(y_i_dat = (pf - x4) / x3,
               x_i_dat =  (ta - x1) / x2 )


```


```{r}

# Create a data frame to use for plotting the UTPC function over its range
dd_utpc <- data.frame(x = seq(-5, 1, length.out = 100))
dd_utpc %<>% mutate(y = (1-x) * exp(x))

# Plot the rescaled data and the UTPC
gg1 <- ggplot(data = combined_dat_params_2, 
              mapping = aes(x = x_i_dat, 
                                       y = y_i_dat)) + 
  geom_point() + 
  facet_wrap(vars(type)) + 
  geom_line(data = dd_utpc,
            mapping = aes(x = x, y = y),
            colour = "red") + 
  # geom_function(fun = function(x) pmax((1-x) * exp(x), 0),
  #               col = "red") + 
  xlim(-6, 2)

print(gg1)
```

## Test on TPCs with negative values

Test on a strictly positive TPC.

```{r}




# extract the first photosynthesis dataset
test1 <- run %>% filter(code == "st")

test1_fit <- do_tpc_fit(test1)


UTPC <- data.frame(x = seq(-6, 1, length.out = 100)) %>%
  mutate(y = (1-x)*exp(x)) %>%
  mutate(ta = (x * test1_fit$x2[1]) + test1_fit$x1[1], 
         pf = (y * test1_fit$x3[1]) + test1_fit$x4[1]
) 

gg_test1 <- ggplot(data = test1_fit, 
                   mapping = aes(x = ta, y = pf)) + 
  geom_point() + 
  geom_line(data = UTPC)

print(gg_test1)

```

Test on the same TPC that is shifted slightly downwards so it includes some negative performance values at either end.

```{r}


# extract the first photosynthesis dataset
test2 <- run %>% filter(code == "st") %>%
  mutate(pf = pf - quantile(pf, probs = 0.25))

test2_fit <- do_tpc_fit(test2)


UTPC <- data.frame(x = seq(-6, 1, length.out = 100)) %>%
  mutate(y = (1-x)*exp(x)) %>%
  mutate(ta = (x * test2_fit$x2[1]) + test2_fit$x1[1], 
         pf = (y * test2_fit$x3[1]) + test2_fit$x4[1]
) 

gg_test2 <- ggplot(data = test2_fit, 
                   mapping = aes(x = ta, y = pf)) + 
  geom_point() + 
  geom_line(data = UTPC)

print(gg_test2)
```

Test on an inverted TPC

```{r}

# extract the first photosynthesis dataset
test3 <- run %>% filter(code == "st") %>%
  mutate(pf = 2 - pf)

test3_fit <- do_tpc_fit(test3)


UTPC <- data.frame(x = seq(-6, 1, length.out = 100)) %>%
  mutate(y = (1-x)*exp(x)) %>%
  mutate(ta = (x * test3_fit$x2[1]) + test3_fit$x1[1], 
         pf = (y * test3_fit$x3[1]) + test3_fit$x4[1]
) 

gg_test3 <- ggplot(data = test3_fit, 
                   mapping = aes(x = ta, y = pf)) + 
  geom_point() 
  # geom_line(data = UTPC)

print(gg_test3)


```


