---
title: "Import data, generate parameter estimates and merge with Rezende data"
output: html_notebook
---


```{r setup}
library(tidyverse)
library(magrittr)
library(readxl)
library(patchwork)
```

This file illustrates on a simple example how a UTPC can be fitted to performance data that contains some negative values. 







### Define our helper functions

These are copied from the main file "import-fit-merge.Rmd". We optimise on the least squares of the rescaled data to the universal thermal performance curve using the L-BFSG-U approach to bound the parameters. 

```{r}

# define our fitness function which is the sum of squares of 
# the predicted - observed
tpc_fit <- function(x, pf, ta){
  
  # extract the parameters
  yopt  <- x[1] # the value of y at the optimum
  Topt  <- x[2] # the optimal temperature
  Delta <- x[3] # the difference between Tc-Topt
  Tc    <- Topt + Delta # calculate Tc
  
  # sum of squares
  sum(((yopt * exp( (ta - Topt) / Delta) * (1 - (ta - Topt)/Delta)) - pf) ^ 2)
  
  
}

# a function to find sensible initial values for the optimiser
get_inits <- function(x){
  
  # initialise the initial parameter value vector
  inits <- c(NA, NA, NA)
  
  # find the largest performance
  idx_max_y <- which.max(x$pf)
  
  # the largest y value is a good guess for yopt
  inits[1] <- x$pf[idx_max_y]
  
  # the corresponding temerature is a good guess for Topt
  inits[2] <- x$ta[idx_max_y]
  
  # use the largest T value as a simple guess for Delta
  inits[3] <-tail(x$ta,1) - inits[2]
  
  # return the initial values
  return(inits)
}


# a wrapper function to do the fit and return a data.frame
# The object x coming in is a data.frame of the same structure
# as the full dataset. The output is a tibble of the same structure
# with columns appended for estimates of the universal TPC with 
# prefix "u_"
do_tpc_fit <- function(x){
  
  # call the optimiser
  fit <- optim(get_inits(x), 
               tpc_fit,
               pf = x$pf, 
               ta = x$ta,
               method = "L-BFGS-B",
               lower = c(0.001, 0.001, 1), 
               upper = Inf)
  
  # return a tibble (data.frame)
  return(x %<>% mutate(u_yopt  = fit$par[1],
                       u_Topt  = fit$par[2],
                       u_Delta = fit$par[3],
                       u_Tc    = fit$par[2] + fit$par[3])
         )

  
}


# this function takes a fitted UTPC model and scales the data to the 
# UTPC. Returns a data.frame the same as the original data.frame
# with new columns for 
# scaled performance pf_utpc 
# and scaled temperature ta_utpc
data2utpc <- function(dd){
  dd %<>% mutate(pf_utpc = pf / u_yopt,
                ta_utpc = (ta - u_Topt) / u_Delta
                )
  return(dd)
}

# This function scales the UTPC curve to the original data based on 
# the fitted UTPC model. Returns a new data.frame with n values of 
# the UTPC function evaluated and rescaled to the data provied in dd.
# Performance can be adjusted according by subtracting the value given in 
# p_adj, and would typically be used to allow for negative performance values
# in the data. 
utpc2data <- function(dd, y_adj = 0, n = 100){
  UTPC <- data.frame(x = seq(-4, 1, length.out = n)) %>%
  mutate(y = (1-x)*exp(x)) %>%
  mutate(ta = (x * dd$u_Delta[1]) + dd$u_Topt[1],
         pf = (y * dd$u_yopt[1] - y_adj)
)
  return(UTPC)
}
```

## Illustration of fitting UTPC to a performance curve with negative values

There are instances where not all performance values are strictly positive. This can occur for example in studies of population growth rates with temperature, where negative values indicate a declining population. The approach that shoudl be taken in these cases is to add an arbitrary amount to the performance values to ensure all data points are strictly positive, fit the UTPC model and then readjust both the data and the UTPC by the same amount. 

```{r}


# This example is the first photosynthesis dataset which we force to have 
# negative performance values by substracting 0.25 from the pf vector.
test2 <- data.frame(ta = c(2.9, 9.7, 15.3, 20.4, 25.5, 
                           30.5, 34.9, 37.3, 39.3),
                    pf = c(0, 0.24, 0.67, 0.93, 1.17, 
                           1.42, 1.49, 0.94, 0) - 0.25)
  # 
  # run %>% filter(code == "st") %>%
  # mutate(pf = pf - quantile(pf, probs = 0.25))

# First we add an arbitrary amount to the performance data to ensure they are 
# positive values. Here, we have looked at the data and decided to artibrarily 
# add 1 to all pf values which is more than enough to ensure they are all 
# positive. Alternatively one could use abs(min(pf)) as an automated way. 
y_adj <- 1

# fit the utpc model after adjusting the performance data, and then 
# transform it back. 
test2_fit <- do_tpc_fit(test2 %>% mutate(pf = pf + y_adj)) %>% 
  mutate(pf = pf - y_adj)

# generate a UTPC on the same scale as the original data.
UTPC <- utpc2data(test2_fit, y_adj = y_adj)

# plot the raw data and the UTPC on the same scale
gg_test2 <- ggplot(data = test2_fit,
                   mapping = aes(x = ta, y = pf)) +
  geom_point() + 
  geom_line(data = UTPC, color = "red")

print(gg_test2)
```

